Storm surge modelling in the Strait of Georgia

### 2014-10-08 RoughWork/manuscript.tex
Added description of statistics and table for Feb 2006

\begin{abstract}
The Strait of Georgia is a large, semi-enclosed body of water between Vancouver Island and the mainland of British Columbia connected to the Pacific Ocean via the Strait of Juan de Fuca at the south and Johnstone Strait at the north. During the winter months, coastal communities in the Strait of Georgia are at risk to flooding caused by storm surges, a natural hazard that can occur when a strong storm coincides with high tide. This investigation produces storm surge hindcasts using a three-dimensional numerical ocean model for the Strait of Georgia and the surrounding bodies of water (Strait of Juan de Fuca, Puget Sound, and Johnstone Strait) collectively known as the Salish Sea. The numerical model employs the Nucleus for European Modelling of the Ocean (NEMO) architecture in a regional configuration. The model is evaluated through comparisons of tidal elevation harmonics, tidal current ellipses, and storm surge water levels with observations. Important forcing factors contributing to storm surges have been assessed, finding that the surge entering the domain from the Pacific Ocean contributes most significantly. In addition, local wind patterns cause a slight increase in surge amplitude on the mainland side of the Strait of Georgia when compared to the Vancouver Island coastal areas. Surge amplitudes are found to be greater within the Strait of Georgia when compared to amplitudes in the Strait of Juan de Fuca. 
%needs a little more work.
\end{abstract}

@@

%storm surge modelling
%murty, danard, crean, non BC models (xu, flather)
Coastal communities can prepare for and respond to storm surge hazards using predictions from storm surge models to determine if flooding conditions are probable. Like coastal ocean models, some of the complexity of a storm surge model depends on the properties of the geographic location of interest. In shallow regions, it is important to include the effects of wetting and drying which allows grid cells to alternate between land and water depending on the modelled sea surface height and depth \citep{hubbert1999storm, weisberg2006hurricane}.  Other regions require modelling over a very large domain since non-local effects from the open sea may influence surge propagation \citep{weisberg2006hurricane, lane2009verification}. Such large domains typically use nested or unstructered grids to resolve the details of complicated coastlines. Appropriate treatment of rivers \citep{flather1994storm}, accurate bathymetry and inlcuding surface waves \citep{xu2010storm} have also been identified as important issues.  Most of these models neglect the effects of stratification, either by treating the ocean as a constant temperature and salinity or by using a two-dimensional depth-averaged set of equations, an assumption that is assumed to be valid.

@@

This article presents the development of a three-dimensional numerical ocean model for the Strait of Georgia and its accuracy in simulating storm surge hindcasts. This baraoclinic model includes the effect of stratification and rivers, however, wetting and drying is not represented. Model evaluation has focused on regions where the near-coast slopes are steep.  This paper is organized as follows: an overview of the model configuration and domain including a description of the ocean model we have employed is provided in section \ref{sec:config}. This section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's accuracy in producing storm surge hindcasts is also assessed in section \ref{sec:storm}. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/tides.pdf}
\caption{Evaluation of model tidal amplitude and phases for the M$_2$ and K$_1$ constituents. On the left, a transect through the stations used for comparison. On the right, modelled M$_2$ (black) and K$_1$ (gray) amplitude (top) and phase (middle) compared with observations in the dashed curves of the same colour. The bottom plot displays the M$_2$ and K_$1$ complex differences.  }
\label{fig:tides}
\end{figure}

\subsection{Stratification} 
%outline: thalweg, plots of winter/summer stratty
%Maybe not this

\section{Storm Surge Hindcasts}\label{sec:storm}
%outline: 
%1. factors contributing to storm surge with focus on Feb 2006 event
%2. Strong wind event but no surge (maybe Dec 2006 or Nov 2009)
%3. High resolution wind (hopefully Dec 2012)
A series of storm surge hindcasts is presented next. Storm surge simulations were started from rest at least three days prior to the event of interest to allow for adjustment of model currents. Stratification conditions from model spin-up at a time close to the event (within 5 days) were chosen to initialize the temperature and salinity fields. Boundary conditions include atmospheric forcing and fresh water input from rivers at the surface, sea surface height anomaly from observations at the open boundary, and tidal forcing with eight tidal constituents. Forcing the model with only eight tidal constituents leads to an error in modelled sea surface height predictions due to the omission of the next leading order tidal constituents such as J$_1$. Comparisons between full tidal predictions and tidal predictions using eight constituents suggest that this error can be up to 40 cm (check) during times of high tidal range, and in particular, during the storm surge season Nov-Feb. As such, storm surge model predictions were corrected by accounting for this error. Additionally, since the sea surface height is calculated about a zero reverence level, the modelled water level is determined by adding a long term average of mean sea level at the specified location to the modelled sea surface height.

Model results have been compared with observations by calculating both the total modelled water level and the modelled residual. The modelled residual is calculated as the difference in model output between a simulation with all forcing conditions and a simulation forced only with tides and rivers. The observed water level is taken from observations by Fisheries and Oceans Canada (citation) and the observed residual is the observations minus tidal predictions. Tidal predictions are calculated using t\_tide \citep{pawlowicz2002classical}, a MATLAB tool that calculates tidal harmonics from a given time series, in this cae, from the year prior surge event on interest. Additionally, several simulations with varying forcing conditions such as removing atmospheric forcing or removing the sea surface height anomaly at the open boundaries have been completed.

@@

@@

On Feb 4, 2006, a strong storm system with winds over 75 km/hr coinciding with an unusually high tide led to significant flooding in coastal communities of the Strait of Georgia \citep{romanowski2010storm}. The maximum water level observed at Point Atkinson on this day was 5.49 m, which is 0.8 m higher than the predicted tides, highlighting the significance of this event as it resulted in a large anomaly and also reached water levels close to the highest level ever recorded (5.61 m). A set of simulations from Feb 1 to 8, 2006 under varying forcing conditions has been completed in order to determine the significance of each factor contributing to the surge.

The model's ability to reproduce this event is demonstrated in Figure \ref{fig:feb2006} which compares observations and corrected model output at Point Atkinson, Victoria, Patricia Bay, and Campbell River. The timing of maximum water level matches well between the observations and the model at all four locations, the most significant contribution to high water level being from the tides. There is also a good match in amplitude of the water level maximum at Point Atkinson and Campbell River, with a relative error less than 4\%. There is only a 7.2 cm difference between observed and modelled water level elevation at Point Atkinson and a 20.9 cm difference at Campbell River. The error at Victoria and Patricia Bay is slightly more significant, reaching 9.6\% and 6.3\% respectively (close to 30 cm in each case), partly due to error in tidal amplitudes as discussed above. The precise location and magnitude of the M$_2$ amphidrome around Victoria is very sensitive to model parameters such as bottom friction.

The error in modelled residuals is also apparent in the panels on the right, most noticeable is an underestimate in maximum residual (or surge) at all locations. There is also a 2-3 hour delay in the timing of the maximum residual except at Victoria where the maximum residual in sync with observations. The errors in maximum residual range from 8.0 cm at Victoria to 28 cm at Campbell River. It is also apparent that Point Atkinson and Campbell River experienced the largest observed surges at 82 cm and 95 cm respectively. The surges at Victoria and Patricia Bay were observed to be 66 cm and 74 cm. 
%Question: is it better to present relative errors or just difference?

@@

@@

\subsection{High Resolution Atmospheric Forcing: Dec 17, 2012}

\subsection{Statistical evaluation}
%I think Hal Ritchie's talk with the storm surges had some stats presented. Look this up...
\citep{bernier2006predicting}

>>>>

\begin{abstract}
The Strait of Georgia is a large, semi-enclosed body of water between Vancouver Island and the mainland of British Columbia connected to the Pacific Ocean via the Strait of Juan de Fuca at the south and Johnstone Strait at the north. During the winter months, coastal communities in the Strait of Georgia are at risk to flooding caused by storm surges, a natural hazard that can occur when a strong storm coincides with high tide. This investigation produces storm surge hindcasts using a three-dimensional numerical ocean model for the Strait of Georgia and the surrounding bodies of water (Strait of Juan de Fuca, Puget Sound, and Johnstone Strait) collectively known as the Salish Sea. The numerical model employs the Nucleus for European Modelling of the Ocean (NEMO) architecture in a regional configuration. The model is evaluated through comparisons of tidal harmonics, tidal current ellipses, and storm surge water levels with observations. Important forcing factors contributing to storm surges have been assessed, finding that the surge entering the domain from the Pacific Ocean contributes most significantly. In addition, local wind patterns cause a slight increase in surge amplitude on the mainland side of the Strait of Georgia when compared to the Vancouver Island coastal areas. Surge amplitudes are found to be greater within the Strait of Georgia when compared to amplitudes in the Strait of Juan de Fuca. 
%needs a little more work.
\end{abstract}

@@

%storm surge modelling
%murty, danard, crean, non BC models (xu, flather)
Coastal communities can prepare for and respond to storm surge hazards using predictions from storm surge models to determine if flooding conditions are probable. Like coastal ocean models, some of the complexity of a storm surge model depends on the properties of the geographic location of interest. In shallow regions, it is important to include the effects of wetting and drying which allows grid cells to alternate between land and water depending on the modelled sea surface height and depth \citep{hubbert1999storm, weisberg2006hurricane}.  Other regions require modelling over a very large domain since non-local effects from the open sea may influence surge propagation \citep{weisberg2006hurricane, lane2009verification}. Such large domains typically use nested or unstructered grids to resolve the details of complicated coastlines. Appropriate treatment of rivers \citep{flather1994storm}, accurate bathymetry and the inclusion surface waves \citep{xu2010storm} have also been identified as important issues.  Most of these models neglect the effects of stratification, either by treating the ocean as a constant temperature and salinity or by using a two-dimensional depth-averaged set of equations, an assumption that is assumed to be valid.

@@

%outline
This article presents the development of a three-dimensional numerical ocean model for the Strait of Georgia and its accuracy in simulating storm surge hindcasts. This baraoclinic model includes the effect of stratification and rivers, however, wetting and drying is not represented. Model evaluation has focused on regions where the near-coast slopes are steep.  This paper is organized as follows: an overview of the model configuration and domain including a description of the ocean model we have employed is provided in section \ref{sec:config}. This section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's accuracy in producing storm surge hindcasts is also assessed in section \ref{sec:storm}. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/tides.pdf}
\caption{Evaluation of model tidal amplitude and phases for the M$_2$ and K$_1$ constituents. On the left, a transect through the stations used for comparison. On the right, modelled M$_2$ (black) and K$_1$ (gray) amplitude (top) and phase (middle) compared with observations in the dashed curves of the same colour. The bottom plot displays the M$_2$ and K$_1$ complex differences.  }
\label{fig:tides}
\end{figure}

\section{Storm Surge Hindcasts}\label{sec:storm}
%outline: 
%1. factors contributing to storm surge with focus on Feb 2006 event
%2. Strong wind event but no surge (maybe Dec 2006 or Nov 2009)
%3. High resolution wind (hopefully Dec 2012)
A series of storm surge hindcasts is presented next. Storm surge simulations were started from rest at least three days prior to the event of interest to allow for adjustment of model currents. Stratification conditions from model spin-up at a time close to the event (within 5 days) were chosen to initialize the temperature and salinity fields. Boundary conditions include atmospheric forcing and fresh water input from rivers at the surface, sea surface height anomaly from observations at the open boundary, and tidal forcing with eight tidal constituents. Forcing the model with only eight tidal constituents leads to an error in modelled sea surface height predictions due to the omission of the next leading order tidal constituents such as J$_1$. Comparisons between full tidal predictions and tidal predictions using eight constituents suggest that this error can be up to 40 cm (check) during times of high tidal range, and in particular, during the storm surge season Nov-Feb. As such, storm surge model predictions were corrected by accounting for this error. Additionally, since the sea surface height is calculated about a zero reference level, the modelled water level is determined by adding a long term average of mean sea level at the specified location to the modelled sea surface height.

Model results have been compared with observations by calculating both the total modelled water level and the modelled residual. The modelled residual is calculated as the difference in model output between a simulation with all forcing conditions and a simulation forced only with tides and rivers. Water level observations are taken from Fisheries and Oceans Canada (citation) and are used to calculate the osberved residual, defined as the observed water level minus tidal predictions. Tidal predictions are calculated using t\_tide \citep{pawlowicz2002classical}, a MATLAB tool that calculates tidal harmonics from a given time series, in this case, from the year prior surge event of interest. Additionally, several simulations with varying forcing conditions such as removing atmospheric forcing or removing the sea surface height anomaly at the open boundaries have been completed.

@@

The model's skill is reported through a few statistical measures of the the total water level $\eta$ and the residual $\delta$. First, the mean absolute error ($MAE$), defined as
\begin{equation}
MAE = \text{mean}\left(\left| M - O \right|\right),
\end{equation}
where $M$ and $O$ are the modelled and observed quantities of interest, is a common measure that estimates the average error in the model. Next, a measure comparing the variance of the error to the variance of the observations defined as \citep{thompson2003prediction},
\begin{equation}
\gamma^2 = \frac{\text{var}\left(M-O\right)}{\text{var}\left(O\right)},
\end{equation}
is also calculated. Smaller values of $\gamma^2$ indicate better model skill. This quantity was used by \citet{bernier2006predicting} and \citet{bernier2010tide} in storm surge predictions in Eastern Canada and is caclulated here as a point of comparison. A third measure, called the Willmott skill score \citep{willmott1982some}, is also used to assess overall model skill in predicting the water level and residual. This quantity is defined as
\begin{equation}
WS = 1 - \frac{\sum_{i=1}^N \left(M_i - O_i\right)^2}{\sum_{i=1}^N \left(|M_i'| + |O_i'|\right)^2},
\end{equation} 
where $M_i' = M_i-\bar{O}$ and $O_i'=O_i-\bar{O}$ and $\bar{O}$ is the mean of the observations. This score is bounded by zero and one. Values close to one indicate good model skill.

@@

On Feb 4, 2006, a strong storm system with winds over 75 km/hr coinciding with an unusually high tide led to significant flooding in coastal communities of the Strait of Georgia \citep{romanowski2010storm}. The maximum water level observed at Point Atkinson on this day was 5.46 m, which is 0.82 m higher than the predicted tides, highlighting the significance of this event as it resulted in a large anomaly and also reached water levels close to the highest level ever recorded (5.61 m). A set of simulations from Feb 1 to 8, 2006 under a variety of forcing conditions has been completed in order to determine the contribution of each forcing factor to the surge.

The model's ability to reproduce this event is demonstrated in Figure \ref{fig:feb2006} which compares observations and corrected model output at Point Atkinson, Victoria, Patricia Bay, and Campbell River. The timing of maximum water level matches well between the observations and the model at all four locations, the most significant contribution to high water level being from the tides which are generally well-represented by the model. There is a good match in the water level maximum at Victoria and Patricia Bay, with a relative error less than 2\%. There is only a 4.2 cm difference between observed and modelled water level elevation at Victoria and a 6.9 cm difference at Patricia Bay, the model slightly underpredicting the maximum water level in each case. The error at Point Atkinson and Campbell River is slightly more significant, reaching 2.5\% and 2.7\% respectively, close to 14 cm overpredicted in each case. It is suprising that Victoria and Patricia Bay agree most closely with observations because the M$_2$ tidal amplitudes in the model are known to be too low in this area. In conjunction with the overpredicted water levels at Point Atkinson and Campbell River, it is likely that one of the forcing factors is overestimated in this example.

A comparison between observed and modelled residuals is also displayed in the panels on the right. In general, there is a very good agreement between the observed and modelled residules. The most noticeable difference is a 4-6 hour delay in the model's timing of the maximum residual except at Victoria where the maximum residual in sync with observations. The errors in maximum residual range from 6.0 cm to high at Victoria to 15.9 cm too low at Campbell River. Point Atkinson and Patricia Bay both experience a very good match with observed residuals with errors less than 2 cm. Note that Point Atkinson and Campbell River observed the largest observed surges at 82 cm and 95 cm respectively. The surges at Victoria and Patricia Bay were observed to be 66 cm and 74 cm.

@@

The maximum water level and maximum residuals are important factors in storm surge hindcasts but a better indicator of overall model performance is provided by a group of statistical measures in Table \ref{tab:feb2006stat}. In general, $\gamma^2$ and $WS$ suggest good model agreement for the predicted water levels, most likely because the tidal signal dominates. The skill scores degrade slightly for the residuals but are still reasonable and indicate very good agreement. Note that the mean absolute error ($MAE$) is lower for the residuals, but this measure does not account for the larger variablity in water levels as opposed to residuals.

\begin{table}[h]
\centering 
\begin{tabular}{l c c c c c c} 
\hline 
& \multicolumn{3}{c}{Water Level ($\eta$)}        & \multicolumn{3}{c}{Residual ($\delta$)} \\ 
\hline 
Location       & $MAE$ (cm) & $\gamma^2$ & $WS$   & $MAE$ (cm) & $\gamma^2$ & $WS$          \\
\hline 
Point Atkinson & 10.5       &   0.012    & 0.994  & 6.2        &  0.0646    & 0.979         \\
Victoria       &  8.6       &   0.032    & 0.992  & 6.1        &  0.0851    & 0.970         \\
Patricia Bay   &  9.1       &   0.024    & 0.992  & 6.9        &  0.0704    & 0.969         \\
Campbell River & 13.1       &   0.041    & 0.990  & 5.8        &  0.0790    & 0.961         \\
\hline 
\end{tabular}
\caption{Evaluation of model performance for the Feb 2006 simulation. Mean absoulte error (MAE), $\gamma^2$, and the Willmott skill score ($WS$) are provided for water level predictions ($\eta$) and resdiuals ($\delta$) during the period displayed in Figure \ref{fig:feb2006}.}
\label{tab:feb2006stat} 
\end{table} 
%anything else to include? statistics calculated in Feb 2006 -paper prodcution.ipynb and saved in statisitcs_feb2006.csv

@@

%\subsection{High Resolution Atmospheric Forcing: Dec 17, 2012}
